Thesis: draft

### 2021-05-24 draft.tex
Update draft.tex
added something in preambule to show numbering of subsubsection
latest update

@@

\textbf{The Watts-Strogatz model in this thesis}\\

@@

\textbf{The stochastic block model in this thesis}\\

@@

Most of the standard opinion dynamics models use a static network in which actors are always active and exchanging opinions. In these models the opinion dynamics might go as follows: each time step a random actor is selected after which its opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all actors at the same time during each time step. \\
\newline
It might be more realistic if actors become active/inactive over time. This would be a better representation of social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the actors gives the network a temporal character. The underlying network structure does not change (each actor has a fixed number of neighbors that does not change over time; no new friends are made or, conversely, no old friends are being defriended). However, by making actors active and inactive the exchange of information between actors becomes time dependent on which actors are active or not. Hence, this does affect the network structure in some way, because if an actor, for example, does not become active for a long period of time, it does not participate in the exchange of information and the network structure behaves like if this actor is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity and clustering may behave differently. \\
\newline
In this thesis, actors are made active with some probability $p_i$ every time step (thus, every time step, each actor becomes active with a probability $p_i$). The activation probability is set equal to $p = 0.1$, which is constant for all the actors. This constant activation probability is chosen for simplicity.

@@

\begin{algorithm}
\caption{Updating the opinions of the active actors at time $t$}
\label{alg:update}
\begin{algorithmic}[1]
	\State \(L_i \leftarrow \text{hidden time-line}\)
	\State \(R_i \leftarrow \text{actual time-line}\)
	\State \(S \leftarrow 20\) \Comment{Maximum length of time-line $R$}
	\State \(r\) \Comment{Stubbornness parameter of each actor, takes values between $0$ and $1$}
	\ForAll{active actor $i$}
		\State \(\text{op} \leftarrow \text{current opinion}\)
		\ForAll{neighbor $k$} \Comment{Add opinion of active actor $i$ to hidden time-line of each neighbor $k$}
			\State \(L_k.\textsc{add}\text{(op)}\)
		\EndFor
		\State \(f \leftarrow \text{rand(0, 1)}\) \Comment{Draw random number between $0$ and $1$}
		\If{$f > r$}
			\State \(\text{op} \leftarrow \Call{Update opinion}{L_i}\) \Comment{Update opinion of active actor $i$}
		\Else
			\State \(\text{op} \leftarrow \text{current opinion}\) \Comment{Do not update opinion of active actor $i$}
		\EndIf
	\EndFor

	\Function{Update opinion}{$L_i$}
		\State \(\textsc{Order}(L_i)\) \Comment{Order hidden time-line according to REF, REC or PR filtering}
		\If{$\text{len}(L_i) > S$}
			\State \(R_i \leftarrow L_i.\textsc{resize}(S)\) \Comment{Time-line $R_i$ = $S$ first elements of ordered hidden time-line}
		\EndIf
		\State \(\text{op} \leftarrow \textsc{probMaj}(R_i)\) \Comment{Update opinion according to probabilistic majority model}
	\State \Return op \Comment{Return updated opinion}
	\EndFunction
\end{algorithmic}
\end{algorithm}
\noindent
Some pseudo-code is given in Algorithm \ref{alg:update} to make the model more comprehensive and understandable. The pseudo-code for updating the opinions of the active actors in case of using a resistance parameter $r$ at time step $t$ is displayed.

@@

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_10x100_01-0001_003-0008_50-50_PR_REC.png}
    \label{sbm_50-50}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10x100_10-001-0001_50-50_PR_REC.png}
    \label{sbm-ws_50-50}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) for balanced starting conditions.]{Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) for balanced starting conditions. \textbf{(a)} SBM; \textbf{(b)} SBM-WS. Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_50_50}
\end{figure}
\newpage
\noindent
\textbf{Echo chambers}\\
\newline
Fig. \ref{echo_50_50} shows the distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$. The region on the $x$-axis close to zero describes neighborhoods in which none, or few actors, adopt opinion B (these are thus neighborhoods in which most actors adopt opinion A). Conversely, the region on the $x$-axis close to one describes neighborhoods in which a majority of the actors adopt opinion B. The distribution is obtained by imposing ten equispaced bins, $\Delta x = 0.1$ \cite{Perra2019}.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_no_stubb_50-50_REC.png}
      \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_no_stubb_50-50_PR_8x7.png}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$, for balanced starting conditions.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$, for balanced starting conditions. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. Four network models are used: (i) ER; (ii) SBM; (iii) SBM-WS and (iv) WS. Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). For the WS model, we use two different values for the rewire probability: $\beta = 0.01$ ($\left<cc\right> = 0.648 \pm 0.003$) and $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). The dotted black line represents a constant amount of echo chambers over time. The distribution is normalized by dividing for the same quantity computed at $t=0$. Each curve is the average of $10^2$ independent simulations. Error bars are omitted for improved visualization.}
\label{echo_50_50}
\end{figure}
\noindent
In Fig. \ref{echo_50_50} two different filtering algorithms are compared, the REC and the PR filtering algorithms. The results for both filtering algorithms are qualitatively the same. The difference is that the formation of echo chambers, using the PR filtering algorithm, is about ten times as large compared to the formation of echo chambers using the REC filtering algorithm. \\
\newline
The ER model shows no, or a limited, sign of echo chambers, for both the REC and PR filtering algorithms. Networks with high clustering (WS and SBM-WS), on the other hand, show a sign of polarization and formation of echo chambers (visible in the two spikes, one around zero and another around one). Going to slightly higher values of the clustering coefficient ($\left<cc\right> = 0.557 \pm 0.007$ to $\left<cc\right> = 0.648 \pm 0.003$) in the WS model results in significantly larger echo chambers. This suggests that the formation of echo chambers is sensitive to changes in the clustering coefficient. The SBM with high modularity, community structure, shows an appearance of polarization (mainly when using the PR filtering algorithm), however, the formation of echo chambers is significantly smaller compared to networks with high clustering. The stochastic block model with low modularity has no, or limited, formation of echo chambers, similar as for the ER model. This is expected, since a SBM with low modularity, and hence, almost non-existing community structure, resembles an ER network. The SBM-WS has both community structure and high average clustering. The formation of echo chambers when using this model is approximately of the same size as the formation of echo chambers in a WS model with similar average clustering. This suggests that it is the high clustering that drives the dynamics to form echo chambers, while the community structure plays a less important role. \\
\newline
For balanced starting conditions, none of the network models, nor the filtering algorithms could break the status quo (Fig. \ref{ev_op_50_50} and related results in \cite{Perra2019}). Despite this preserving of the balanced opinion distribution, the interplay of filtering algorithms and network structure can potentially shift the initial randomly distributed opinions to a clustered opinion distribution and thus lead to polarization and formation of echo chambers.

@@

We will now investigate the effect of unbalanced starting conditions, that is one opinion is initially more prevalent than the other. The initial prevalence of opinion A is $P_\text{A}(0) = 0.2$, the minority opinion, and of opinion B is $P_\text{B}(0) = 0.8$, the majority opinion.\\ 
\newline
\textbf{Evolution of the prevalence of the opinions}\\
\newline
Fig. \ref{ev_op_20_80} shows the results for the evolution of the prevalence of opinion A, which is the minority opinion, over time. For each network model the REC and PR filtering algorithms are compared. \\
\newline
The REC filtering algorithm maintains the status quo; the prevalence of opinion A remains stable over time for the four network models. When using the PR filtering algorithm, results are different. The prevalence of the minority opinion, $P_\text{A}(t)$, decreases over time. This decrease is a consequence of the PR filtering algorithm. Since the PR filtering algorithm orders the posts according to the actors' preferences, actors with the majority opinion B will mainly see posts with the same opinion B on their time-line, which limits the chance of changing their opinion to the minority opinion. Actors with the minority opinion A, on the other hand, will see posts with opinion A on their time-line when using the PR filtering algorithm. However, because there are more actors with opinion B than actors with opinion A, the actors with the minority opinion A will also see a significant number of posts with the majority opinion B on their time-line. When using the PR filtering algorithm, there is an imbalance in the sense that actors with the minority opinion A have a chance of switching to the majority opinion B, whereas actors with the majority opinion B have (almost) no chance of switching to the minority opinion A. This imbalance leads to the decrease of the minority opinion A.

\begin{figure}[H]
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_ER_001_50-50_PR_REC_20-80_8x7.png}
      \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_10-006_PR_REC_20-80_8x7.png}
    \label{ws_20-80_op}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}    
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_01-0001_003-0008_PR_REC_20-80_8x7.png}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10-001-0001_PR_REC_20-80_8x7.png}
    \label{sbm-ws_20-80_op}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions.]{Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. \textbf{(a)} ER; \textbf{(b)} WS; \textbf{(c)} SBM; \textbf{(d)} SBM-WS. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_20_80}
\end{figure}
\noindent
Fig. \ref{ev_op_20_80} shows that the minority opinion A eventually vanishes. This is the case for the low clustering networks, with $P_\text{A}(500) = 0.001 \pm 0.003$ for the ER network, $P_\text{A}(500) = 0.002 \pm 0.003$ for the SBM with a low modularity and $P_\text{A}(500) = 0.003 \pm 0.006$ for the SBM with a high modularity. The high clustering in the WS and SBM-WS networks seems to slightly hamper this vanishing of the minority opinion with an average value of $P_\text{A}(500) = 0.007 \pm 0.014$ for the SBM-WS network (Fig. \ref{sbm-ws_20-80_op}) and an average value of $P_\text{A}(500) = 0.009 \pm 0.015$ for the WS network (Fig. \ref{ws_20-80_op}).\\ 
\newline
This result differs from results in \cite{Perra2019}, where the minority opinion was able to survive over time for all the network models, even for the PR filtering algorithm, and where a minority opinion prevalence at $t=500$ between $P_\text{A}(500) = 0.06$ and $P_\text{A}(500) = 0.09$ was reported \cite{Perra2019}. The difference in the results is most likely due to an interplay between the network size $N$, the average degree $\left<k\right>$ and the maximum length of the actual time-line $S$. The results that where obtained in \cite{Perra2019} are for networks with size $N = 10^5$ nodes and average degree $\left<k\right> \sim 6$. In this thesis results are obtained for $N=10^3$ and average degree $\left<k\right> \sim 10$. In both cases the maximum length of the actual time-line is $S=20$. Fig. \ref{ev_op_WS_k=6} shows the results for the evolution of the prevalence of opinion A for an initial $20/80$ opinion distribution and using the PR filtering algorithm; Fig. \ref{ws1000} displays results for a WS model with $N=10^3$ and $\left<k\right> = 6$; Fig. \ref{ws10000} shows results for a WS model with $N=10^4$ and $\left<k\right> = 6$. We see that the minority opinion A reaches higher final values of $P_0(500)$ compared to the case depicted in Fig. \ref{ws_20-80_op}, with a value of $P_\text{A}(500) = 0.027 \pm 0.023$ in Fig. \ref{ws1000} and a value of $P_\text{A}(500) = 0.030 \pm 0.018$ in Fig. \ref{ws10000}.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_1000-6_PR_20-80_8x7.png}
    \label{ws1000}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_10000-6_PR_20-80_8x7.png}
    \label{ws10000}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. A WS model with $N=10^3$ or $N=10^4$, $\left<k\right>=6$ and $\beta = 0.01$ is implemented. The PR filtering algorithm is used.]{Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. The Watts-Strogatz network with $\left<k\right> = 6$ and $\beta = 0.01$ is depicted with \textbf{(a)} $N=10^3$; \textbf{(b)} $N=10^4$. In both networks the average clustering coefficient is $\left<cc\right> = 0.584 \pm 0.003$. The PR filtering filtering algorithm is used. The dotted black represents the initial prevalence of opinion A. The dotted grey line represents a vanished prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_WS_k=6}
\end{figure}
\noindent
\textbf{Echo chambers}\\
\newline
Fig. \ref{echo_20_80} shows the distribution of the fraction of friends of actor $i$ with the same opinion B at $t=500$.  The REC and PR filtering algorithms are compared for the ER, WS, SBM and SBM-WS.\\
\newline
The results for the ER model and the SBM (both the high and low modularity configuration) look qualitatively the same. Similarly, the results for the WS model and the SBM-WS agree with each other as well. This difference in results between ER/SBM (negligible clustering) and WS/SBM-WS (high clustering) suggests that it is the clustering that drives these differences.

@@

There are no visible effects of having a community structure. Fig. \ref{sbm} shows the results for a SBM with high modularity (community structure) and low modularity (negligible community structure). There is (almost) no difference in the results for high and low modularity.\\

@@

Lets now study the situation for unbalanced starting conditions. The initial prevalence of opinion A is $P_\text{A}(0) = 0.2$ and of opinion B is $P_\text{B}(0) = 0.8$.

@@

>>>>

\setcounter{tocdepth}{3}
\setcounter{secnumdepth}{3}

@@

\textbf{Watts-Strogatz model in this thesis}\\

@@

\textbf{Stochastic block model in this thesis}\\

@@

Most of the standard opinion dynamics models use a static network in which actors are always active and exchanging opinions. In these models the opinion dynamics might go as follows: each time step a random actor is selected after which its opinion is updated according to the opinion dynamics model that is implemented. Another possibility is updating all actors at the same time during each time step. \\
\newline
It might be more realistic if actors become active/inactive over time. This would be a better representation of social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the actors gives the network a temporal character. The underlying network structure does not change (each actor has a fixed number of neighbors that does not change over time; no new friends are made or, conversely, no old friends are being unfriended). However, by making actors active and inactive the exchange of information between actors becomes time dependent on which actors are active or not. Hence, this does affect the network structure in some way, because if an actor, for example, does not become active for a long period of time, it does not participate in the exchange of information and the network structure behaves like if this actor is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity and clustering may behave differently. \\
\newline
In this thesis, actors are made active with some probability $p_i$ every time step (thus, every time step, each actor becomes active with a probability $p_i$). The activation probability is set equal to $p = 0.1$, which is constant for all the actors. This constant activation probability is chosen for simplicity.

@@

\begin{algorithm}[H]
\caption{Updating the opinions of the active actors at time $t$}
\label{alg:update}
\begin{algorithmic}[1]
	\State \(L_i \leftarrow \text{hidden time-line}\)
	\State \(R_i \leftarrow \text{actual time-line}\)
	\State \(S \leftarrow 20\) \Comment{Maximum length of time-line $R$}
	\State \(r\) \Comment{Stubbornness parameter of each actor, takes values between $0$ and $1$}
	\ForAll{active actor $i$}
		\State \(\text{op} \leftarrow \text{current opinion}\)
		\ForAll{neighbor $k$} \Comment{Add opinion of active actor $i$ to hidden time-line of each neighbor $k$}
			\State \(L_k.\textsc{add}\text{(op)}\)
		\EndFor
		\State \(f \leftarrow \text{rand(0, 1)}\) \Comment{Draw random number between $0$ and $1$}
		\If{$f > r$}
			\State \(\text{op} \leftarrow \Call{Update opinion}{L_i}\) \Comment{Update opinion of active actor $i$}
		\Else
			\State \(\text{op} \leftarrow \text{current opinion}\) \Comment{Do not update opinion of active actor $i$}
		\EndIf
	\EndFor

	\Function{Update opinion}{$L_i$}
		\State \(\textsc{Order}(L_i)\) \Comment{Order hidden time-line according to REF, REC or PR filtering}
		\If{$\text{len}(L_i) > S$}
			\State \(R_i \leftarrow L_i.\textsc{resize}(S)\) \Comment{Time-line $R_i$ = $S$ first elements of ordered hidden time-line}
		\EndIf
		\State \(\text{op} \leftarrow \textsc{probMaj}(R_i)\) \Comment{Update opinion according to probabilistic majority model}
	\State \Return op \Comment{Return updated opinion}
	\EndFunction
\end{algorithmic}
\end{algorithm}
\noindent
Pseudo-code is given in Algorithm \ref{alg:update} to make the model more comprehensive and understandable. The pseudo-code for updating the opinions of the active actors in case of using a resistance parameter $r$ at time step $t$ is displayed.

@@

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_10x100_01-0001_003-0008_50-50_PR_REC_8x7.png}
    \label{sbm_50-50}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10x100_10-001-0001_50-50_PR_REC.png}
    \label{sbm-ws_50-50}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) for balanced starting conditions.]{Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) for balanced starting conditions. \textbf{(a)} SBM; \textbf{(b)} SBM-WS. Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_50_50}
\end{figure}
\noindent
\textbf{Echo chambers}\\
\newline
Fig. \ref{echo_50_50} shows the distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$. The region on the $x$-axis close to zero describes neighborhoods in which none, or few actors, adopt opinion B (these are thus neighborhoods in which most actors adopt opinion A). Conversely, the region on the $x$-axis close to one describes neighborhoods in which a majority of the actors adopt opinion B. The distribution is obtained by imposing ten equispaced bins, $\Delta x = 0.1$ \cite{Perra2019}.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_no_stubb_50-50_REC_8x7.png}
      \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_no_stubb_50-50_PR_8x7.png}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$, for balanced starting conditions.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$, for balanced starting conditions. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. Four network models are used: (i) ER; (ii) SBM; (iii) SBM-WS and (iv) WS. Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). For the WS model, we use two different values for the rewire probability: $\beta = 0.01$ ($\left<cc\right> = 0.648 \pm 0.003$; WS2) and $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$; WS1). The dotted black line represents a constant amount of echo chambers over time. The distribution is normalized by dividing for the same quantity computed at $t=0$. Each curve is the average of $10^2$ independent simulations. Error bars are omitted for improved visualization.}
\label{echo_50_50}
\end{figure}
\noindent
In Fig. \ref{echo_50_50} two different filtering algorithms are compared, the REC and the PR filtering algorithms. The results for both filtering algorithms are qualitatively the same. The difference is that the formation of echo chambers, using the PR filtering algorithm, is about ten times as large compared to the formation of echo chambers using the REC filtering algorithm. \\
\newline
The ER model shows no, or a limited, sign of echo chambers, for both the REC and PR filtering algorithms. Networks with high clustering (WS and SBM-WS), on the other hand, show a sign of polarization and formation of echo chambers (visible in the two spikes, one around zero and another around one). Going to slightly higher values of the clustering coefficient ($\left<cc\right> = 0.557 \pm 0.007$ to $\left<cc\right> = 0.648 \pm 0.003$) in the WS model results in significantly larger echo chambers. This suggests that the formation of echo chambers is sensitive to changes in the clustering coefficient. The SBM with high modularity, community structure, shows an appearance of polarization (mainly when using the PR filtering algorithm), however, the formation of echo chambers is significantly smaller compared to networks with high clustering. The stochastic block model with low modularity has no, or limited, formation of echo chambers, similar as for the ER model. This is expected, since a SBM with low modularity, and hence, almost non-existing community structure, resembles an ER network. The SBM-WS has both community structure and high average clustering. The formation of echo chambers when using this model is approximately of the same size as the formation of echo chambers in a WS model with similar average clustering. This suggests that it is the high clustering that drives the dynamics to form echo chambers, while the community structure plays a less important role. \\
\newline
For balanced starting conditions, none of the network models, nor the filtering algorithms could break the status quo (Fig. \ref{ev_op_50_50} and related results in \cite{Perra2019}). Despite this preserving of the balanced opinion distribution, the interplay of filtering algorithms and network structure can potentially shift the initial randomly distributed opinions to a clustered opinion distribution and thus lead to polarization and formation of echo chambers.

@@

We will now investigate the effect of unbalanced starting conditions, that is one opinion is initially more prevalent than the other. The initial prevalence of opinion A is $P_\text{A}(0) = 0.2$, the minority opinion, and of opinion B is $P_\text{B}(0) = 0.8$, the majority opinion.\\ 
\newline
\textbf{Evolution of the prevalence of the opinions}\\
\newline
Fig. \ref{ev_op_20_80} shows the results for the evolution of the prevalence of opinion A, which is the minority opinion, over time. For each network model the REC and PR filtering algorithms are compared. \\
\newline
The REC filtering algorithm maintains the status quo; the prevalence of opinion A remains stable over time for the four network models. When using the PR filtering algorithm, results are different. The prevalence of the minority opinion, $P_\text{A}(t)$, decreases over time. This decrease is a consequence of the PR filtering algorithm. Since the PR filtering algorithm orders the posts according to the actors' preferences, actors with the majority opinion B will mainly see posts with the same opinion B on their time-line, which limits the chance of changing their opinion to the minority opinion. Actors with the minority opinion A, on the other hand, will see posts with opinion A on their time-line when using the PR filtering algorithm. However, because there are more actors with opinion B than actors with opinion A, the actors with the minority opinion A will also see a significant number of posts with the majority opinion B on their time-line. When using the PR filtering algorithm, there is an imbalance in the sense that actors with the minority opinion A have a chance of switching to the majority opinion B, whereas actors with the majority opinion B have (almost) no chance of switching to the minority opinion A. This imbalance leads to the decrease of the minority opinion A.\\
\newline
Fig. \ref{ev_op_20_80} shows that the minority opinion A eventually vanishes. This is the case for the low clustering networks, with $P_\text{A}(500) = 0.001 \pm 0.003$ for the ER network, $P_\text{A}(500) = 0.002 \pm 0.003$ for the SBM with a low modularity and $P_\text{A}(500) = 0.003 \pm 0.006$ for the SBM with a high modularity. The high clustering in the WS and SBM-WS networks seems to slightly hamper this vanishing of the minority opinion with an average value of $P_\text{A}(500) = 0.007 \pm 0.014$ for the SBM-WS network (Fig. \ref{sbm-ws_20-80_op}) and an average value of $P_\text{A}(500) = 0.009 \pm 0.015$ for the WS network (Fig. \ref{ws_20-80_op}).

\begin{figure}[H]
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_ER_001_50-50_PR_REC_20-80_8x7.png}
      \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_10-006_PR_REC_20-80_8x7.png}
    \label{ws_20-80_op}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}    
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_01-0001_003-0008_PR_REC_20-80_8x7.png}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10-001-0001_PR_REC_20-80_8x7.png}
    \label{sbm-ws_20-80_op}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions.]{Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. \textbf{(a)} ER; \textbf{(b)} WS; \textbf{(c)} SBM; \textbf{(d)} SBM-WS. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_20_80}
\end{figure}
\noindent
This result differs from results in \cite{Perra2019}, where the minority opinion was able to survive over time for all the network models, even for the PR filtering algorithm, and where a minority opinion prevalence at $t=500$ between $P_\text{A}(500) = 0.06$ and $P_\text{A}(500) = 0.09$ was reported \cite{Perra2019}. The difference in the results is most likely due to an interplay between the network size $N$, the average degree $\left<k\right>$ and the maximum length of the actual time-line $S$. The results that where obtained in \cite{Perra2019} are for networks with size $N = 10^5$ nodes and average degree $\left<k\right> \sim 6$. In this thesis results are obtained for $N=10^3$ and average degree $\left<k\right> \sim 10$. In both cases the maximum length of the actual time-line is $S=20$. Fig. \ref{ev_op_WS_k=6} shows the results for the evolution of the prevalence of opinion A for an initial $20/80$ opinion distribution and using the PR filtering algorithm; Fig. \ref{ws1000} displays results for a WS model with $N=10^3$ and $\left<k\right> = 6$; Fig. \ref{ws10000} shows results for a WS model with $N=10^4$ and $\left<k\right> = 6$. We see that the minority opinion A reaches higher final values of $P_0(500)$ compared to the case depicted in Fig. \ref{ws_20-80_op}, with a value of $P_\text{A}(500) = 0.027 \pm 0.023$ in Fig. \ref{ws1000} and a value of $P_\text{A}(500) = 0.030 \pm 0.018$ in Fig. \ref{ws10000}.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_1000-6_PR_20-80_8x7.png}
    \label{ws1000}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_10000-6_PR_20-80_8x7.png}
    \label{ws10000}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. A WS model with $N=10^3$ or $N=10^4$, $\left<k\right>=6$ and $\beta = 0.01$ is implemented. The PR filtering algorithm is used.]{Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. The Watts-Strogatz network with $\left<k\right> = 6$ and $\beta = 0.01$ is depicted with \textbf{(a)} $N=10^3$; \textbf{(b)} $N=10^4$. In both networks the average clustering coefficient is $\left<cc\right> = 0.584 \pm 0.003$. The PR filtering filtering algorithm is used. The dotted black represents the initial prevalence of opinion A. The dotted grey line represents a vanished prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_WS_k=6}
\end{figure}
\noindent
\textbf{Echo chambers}\\
\newline
Fig. \ref{echo_20_80} shows the distribution of the fraction of friends of actor $i$ with the same opinion B at $t=500$.  The REC and PR filtering algorithms are compared for the ER, WS, SBM and SBM-WS.\\
\newline
The results for the ER model and the SBM (both the high and low modularity configuration) look qualitatively the same. Similarly, the results for the WS model and the SBM-WS agree with each other as well. This difference in results between ER/SBM (negligible clustering) and WS/SBM-WS (high clustering) suggests that it is the clustering that drives these differences.\\
\newline
There are no visible effects of having a community structure. Fig. \ref{sbm} shows the results for a SBM with high modularity (community structure) and low modularity (negligible community structure). There is (almost) no difference in the results for high and low modularity.

@@

@@

Lets now study the situation for unbalanced starting conditions. The initial prevalence of opinion A is $P_\text{A}(0) = 0.2$ and of opinion B is $P_\text{B}(0) = 0.8$.\\
In Fig. \ref{echo_vs_frac_complStubb_PR-REC_20-80}, the formation of echo chambers versus the fraction of stubborn actors, $f_s$ is shown for the PR and the REC filtering algorithms. The interpretation of the $y$-axis is different than for configurations with balanced starting conditions. For balanced starting conditions, the echo chamber size was interpreted as the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. We first normalized the echo chambers around opinion A and those around opinion B and then took the average of both. For unbalanced starting conditions, we have skewed echo chamber distributions, as can be seen in Fig. \ref{echo_20_80}, with normalized values going to infinity. We still want to have an average of the echo chambers around opinion A and those around opinion B to quantify the total formation of echo chambers, but we want to avoid problems with those infinities. Hence, we opted for a different average: first we calculate the total number of like-minded neighborhoods at $t=500$ (the sum of the number of echo chambers around opinion A and those around opinion B), then we divide that quantity by the total number of like-minded neighborhoods at $t=0$ (we first take the average and than normalize it, instead of normalizing first and then taking the average, as was done for balanced starting conditions). By taking the average, we neglect the skewed echo chamber distribution. This is not a problem. We want a sense of the total number of echo chambers in the system and how this total number is impacted by the presence of stubborn actors and are less interested in the distribution of the echo chambers.  It is however not recommended to (quantitatively) compare the formation of echo chambers between the different network models when depicting the results as an average in case of unbalanced starting conditions. When comparing the different network models, the skewed distributions should not be overlooked. The goal of this figure is to show the dependency of the formation of echo chambers in the network models on the presence of a fraction of stubborn actors and not to (quantitatively) compare the different models against each other.

@@

  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_comm05_distr_REC_8x7.png}
    \label{rec_comm05_echo}
